Model update for: Linking pinyon jay population abundance to pinyon cone production and climate in the SW United States

### 2025-04-29 drafts/draftOct2024ebd.qmd
updating code for model to get oos data prepped and also RMSE for both in and out of sample

Pinyon jays are a widespread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico). Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with pinyon pines (CITE Balda 2020, Ligon 1978). Birds cache large amounts of pinyon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982). Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies). Pinyon jays are in decline across much of their rangefor a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen). Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].

Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS). \[Any other justifications for including only data from this region WRT climate/habitat? Please add tree guys\].

@@

eBIRD is community science data, meaning that non-scientist birders collect eBIRD data and contribute to the eBIRD database. We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE). Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest). We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols. We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (CITE). We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.

eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling [@robinson2018]. To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates. We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space. Given that we observed the fewest number of checklists per year in 2010 (4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell). After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them. Thus, we further reduced temporal bias by subsampling checklists again so that we had relatively equal sample sizes across years based on the year with the fewest checklists (n = 1715 checklists in 2010). We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production. As a result, we had, on average \_\_ checklists per grid cell in a specific year.

@@

To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE). The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions). The biological process includes a likelihood that describes the biological drivers of latent abundance. These covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].

@@

@@

| Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
|----------------|----------------------------|-----------------------------|
| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?                    |
| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                                | How is pinyon jay abundance linked to physiological limits and/or insect availability?                                            |
| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                            | How is pinyon jay abundance linked to insect availability?                                                                        |
| Monsoonality                       | The degree to which the site depends on monsoon moisture (percent of total annual precipitation occurring in July and August) | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?                                         |
| Pinyon basal area                  | The basal area of pinyon pine                                                                                                 | How is pinyon jay abundance shaped by local habitat preferences for food trees?                                                   |
| Cone abundance x Temperature       |                                                                                                                               | Is the importance of cone abundance for jays mediated by temperature?                                                             |
| Cone abundance x Precipiation      |                                                                                                                               | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
| Cone abundance x Monsoonality      |                                                                                                                               | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?        |
| Cone abundance x Pinyon basal area |                                                                                                                               | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine? |

@@

>>>>

Pinyon jays are a widespread species that occurs across the western US (California, Nevada, Utah, Arizona, Colorado, New Mexico). Pinyon jays are generalist seed- and insect-eaters and are part of an important mutualistic relationship with pinyon pines (CITE Balda 2020, Ligon 1978). Birds cache large amounts of pinyon pine seeds when they are available and by burying seeds they increase the potential for seedlings to germinate (CITE: Floyd 1982). Pinyon jays often flock and nest in large numbers and are highly social \[maybe examples\] (CITE some behavior studies). Pinyon jays are in decline across much of their range for a variety of potential reasons, including changes in habitat quality and food availability (CITE Boone, VanLannen). Concurrently, pinyon pines have seen shifts in their range and reproduction \[Andreas - provide some details here about decline/change for pinyon pine\].

Although pinyon jays occur across the western US, in this study, we focused on the intermountain and southwest regions of the US because of available information on cone production for single-needle pinyon (e.g., data from Utah, Colorado, New Mexico, and Arizona; CITE ANDREAS). \[Any other justifications for including only data from this region WRT climate/habitat? Please add tree guys\].

@@

eBIRD is community science data, meaning that non-scientist birders collect bird observation data and contribute to the eBIRD database. We followed filtering methods from the eBIRD database curators (Cornell Lab CITE/LINK HERE). Specifically, we selected data that represented only complete checklists (e.g., observers recorded all birds of all species observed, not just individuals of species of interest). We selected checklists from the nesting season (late February to early May CITE) and included only observations of "stationary" and "traveling" protocols. We considered only the nesting season because this is a time of year in which the mechanisms for local bird abundance are easy to evaluate because of communal nesting behavior and because nesting often occurs near areas where birds have cached seeds from the previous year (CITE). We only included checklists that were collected in under five hours, 0-5 km in distance, and with ten or fewer observers.

eBIRD data is unstructured data without a regular sampling design, so we performed spatial and temporal sub-sampling [@robinson2018]. To do this, we divided the landscape into 4x4 km grid cells to match the spatial scale of the covariates. We spatially sub-sampled eBIRD observations to include a set of checklists in each grid cell that represented relatively equal sampling effort across time and space. Given that we observed the fewest number of checklists per year in 2010 (average 4.2 checklists per grid cell), we subsampled to this level (4 checklists per cell). After filtering to this level, we still had fewer checklists in early years due to fewer cells having checklists in them. Thus, we further reduced temporal bias by subsampling checklists again so that we had relatively equal sample sizes across years based on the year with the fewest checklists (n = 1715 checklists in 2010). We performed this temporal sampling randomly (e.g., not by 4x4 km grid cell) We also removed all checklists within 10 km of the exterior edges of the four states for which we had data (Arizona, Colorado, New Mexico, and Utah) since we had masked some other covariate layers to these states and observations near edges may have misrepresented values for pinyon basal area and cone production. As a result, we had, on average \_\_ checklists per grid cell in a specific year.

@@

To explore relationships between pinyon jay abundance and environmental covariates, we used a two-part model with an observation process for eBIRD and a biological process model that estimates latent "abundance" values (CITE). The observation process models include covariates that describe variation in detection probabilities (e.g., survey conditions). The biological process includes a likelihood that describes the biological drivers of latent abundance. These biological covariates include a stochastic antecedent modeling structure that allows cone abundance to have both lead and lagged effects (e.g., bird abundance can respond before or after cone maturation) and other yearly climate variables to have lagged effects [@ogle2015].

@@

<!--# Fix this section for appropriate calculation and indexing of N! waiting on models to know what this should be... -->

@@

| Covariate or interaction           | Description                                                                                                                   | Question                                                                                                                          |
|------------------------------------|-------------------------------------------------------------------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------|
| Pinyon pine cone abundance         | Abundance of cones from a predictive model (CITE)                                                                             | How is pinyon jay abundance linked to pinyon cone availability (can have lead, immediate, and lagged effects)?                    |
| Maximum temperature                | A proxy for physiological limits and/or insect resource availability (can have lagged effects)                                | How is pinyon jay abundance linked to physiological limits and/or insect availability?                                            |
| Cumulative precipitation           | A proxy for insect resource availability (can have lagged effects)                                                            | How is pinyon jay abundance linked to insect availability?                                                                        |
| Monsoonality                       | The degree to which the site depends on monsoon moisture (percent of total annual precipitation occurring in July and August) | How is pinyon jay abundance shaped by broad-scale climatic mediators of cone reliability?                                         |
| Pinyon basal area                  | The basal area of pinyon pine                                                                                                 | How is pinyon jay abundance shaped by local habitat preferences for food trees?                                                   |
| Cone abundance x Temperature       |                                                                                                                               | Is the importance of cone abundance for jays mediated by temperature?                                                             |
| Cone abundance x Precipiation      |                                                                                                                               | Is the importance of cone abundance for jays mediated by precipitation (e.g., more/less availability of insect prey)              |
| Cone abundance x Monsoonality      |                                                                                                                               | Is the importance of cone abundance for jays different across a range of monsoonal conditions that alter cone reliability?        |
| Cone abundance x Pinyon basal area |                                                                                                                               | Is the importance of cone abundance for jays different depending on the local habitat configuration and abundance of pinyon pine? |

@@

### 2.6 Out of sample model validation

To evaluate model fit, we performed the same set of steps for spatially and temporally stratifying available checklists from eBIRD (section \_\_ above), ensuring that we selected checklists that were not used in model fitting. We used model results (median values for covariate effects and intercepts) to calculate estimates for this dataset and used RMSE to evaluate model fit.

<!--# From Stack Overflow: "RMSE has the same unit as the dependent variable (DV). It means that there is no absolute good or bad threshold, however you can define it based on your DV. For a datum which ranges from 0 to 1000, an RMSE of 0.7 is small, but if the range goes from 0 to 1, it is not that small anymore. However, although the smaller the RMSE, the better, you can make theoretical claims on levels of the RMSE by knowing what is expected from your DV in your field of research. Keep in mind that you can always normalize the RMSE." The RMSE for your training and your test sets should be very similar if you have built a good model. If the RMSE for the test set is much higher than that of the training set, it is likely that you've badly over fit the data, i.e. you've created a model that tests well in sample, but has little predictive value when tested out of sample. -->

### 2025-04-29 meetings/april2025_updates.qmd
updating code for model to get oos data prepped and also RMSE for both in and out of sample

source(here('code',
       '03_visualization',
       'model_betas_lags.R'))

@@

I am aiming for a general framing around co-evolution and conservation given changes in cone availability over the last century. I'm currently looking at Journal of Animal Ecology as a potential outlet (short format - 8500 words inclusive of all text, including references).

@@

-   cone abundance (with yearly lead, immediate, and lag effects)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Temporal hypotheses for cone-jay abundance relationship. Birds could respond in a combination of time periods, including before, concurrent with, or after mature cones are available."

@@

We spatially and temporally stratified checklists in our analyses based on an adaption of the [best practices](https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html) outlined by Cornell Labs for using eBIRD data for science. We spatially sub-sampled based on a 4x4 km grid (since this was the scale of all our covariates). We incorporated spatial uncertainty by then buffering each checklist within that grid by 1/2 of its sampling distance from eBIRD and then merging all checklists in a 4x4 km grid into a "blob" that represented the spatial unit for estimating jay abundance (checklists = repeat observations within a site/year). Within the blob, we determined the amount of the blob overlapping the original 4x4 km grid cell and all surrounding cells and used covariates calculated as the weighted average of covariate values in each cell times the amount of overlap of the blob with that cell. Further, we multiplied the rate parameter, $\lambda$, in the Poisson distribution estimating latent *N* by the area of the blob (since blobs represented different sampling areas).

You can examine the model code [here](https://github.com/anamtk/pinyon_jays/blob/main/code/jags_models/ebird_abund_JAGS_nospuncert2.R).

@@

Here are the overall results for covariate effects, with all covariates scaled so relative importance is comparable.

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Temporal effect of cone abundance effect on pinyon jays. Jays respond in abundance the same year as cone initiation (year -2) and also in the year following cone maturation and three years following cone maturation."

@@

-   Birds increase in abundance the same year that cones initiate. Something about cone climate triggers also seem to help birds?

@@

To note: will have to think about these temporal signals and make sure we're on the right track. Cones are available in the late summer of all the years; all eBIRD data is coming from the nesting season of that year (February - April).

@@

## Next steps

Take a gander at these results, think about them in the context of your knowledge areas.

General thoughts/feedback:

-   I'm keen to keep this covariate and interaction set since we're really interested in cone effect mediation. If you feel very strongly that this is not sufficient, happy to hear the case for adding something.

-   Is there anything glaringly wrong/missing from the model? (Jamie and Kiona)

-   What are your hot takes on the relationships with covariates and their biological reasons?

I will aim to schedule a collaborator meeting on this project in the weeks after spring break (after next week) and look forward to chatting with folks and moving this project towards the door!

>>>>

source(here('code',
       '03_visualization',
       'weight_corrected_covarate_effects.R'))

@@

@@

-   cone abundance (with yearly lead and lag effects)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 1: Temporal hypotheses for cone-jay abundance relationship. Birds could respond in a combination of time periods, including before or after mature cones are available."

@@

We spatially and temporally stratified checklists in our analyses based on an adaption of the [best practices](https://cornelllabofornithology.github.io/ebird-best-practices/abundance.html) outlined by Cornell Labs for using eBIRD data for science. We spatially sub-sampled based on a 4x4 km grid (since this was the scale of all our covariates). We incorporated spatial uncertainty by then buffering each checklist within that grid by 1/2 of its sampling distance from eBIRD and then merging all checklists in a 4x4 km grid into a "blob" that represented the spatial unit for estimating jay relationships with covariates and expected abundance, $\lambda$ (checklists = repeat observations within a site/year). Within the blob, we determined the amount of the blob overlapping the original 4x4 km grid cell and all surrounding cells and used covariates calculated as the weighted average of covariate values in each cell times the amount of overlap of the blob with that cell. For every checklist in a blob, we determined "true" abundance, *N*, by multiplying the blob-level rate parameter, $\lambda$, in the Poisson distribution by the area surveyed by each checklist in that blob. Thus, *N* is estimated at the checklist level and shares biological information across checklists within a blob.

You can examine the model code [here](https://github.com/anamtk/pinyon_jays/blob/main/code/jags_models/ebird_abund_JAGS_checkN_KO.R).

@@

### Main effects of covariates

Here are the overall results for covariate effects (I know they're missing 95% BCI - have to calculate after the fact because reasons, and haven't gotten to it yet! All are non-zero except maybe ConexBasal Area interaction). All are scaled and so their magnitudes can be compared.

@@

### Cone temporal effects

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Figure 3: Temporal effect of cone abundance effect on pinyon jays. Jays respond in abundance the spring before summer/fall cone initiation (2 years before cones mature) and also in the year following cone maturation and three years following cone maturation."

@@

-   Birds increase in abundance the spring before cones initiate in the fall/summer. Something about climate triggers also seem to help birds?

@@

### Climate temporal effects

@@

### Interactions

I haven't had a chance to spend too much time in the interactions, beyond what I've highlighted above about their main effects. However, these also include temporal effects, so they have sort of a "this, then this, then this" type relationship with cones and birds. For example, the positive interaction between cones and precipitation means that birds rely more on cones in wetter periods. Two of the most important wet periods are the breeding and fledging seasons (spring/early summer) two years ago. Thus, when there is a wet spring/early summer followed by a cone initiation year, there will be more birds in areas where cones are maturing both in that year and in the year following cone maturation (2 years later). Maybe it's more important to focus just on the main effects rather than get bogged down in the temporal aspects of the interactions... not sure. But given that wet breeding seasons 2 years ago lead to fewer birds and that birds rely more on cones in wet times, maybe the more important factor here is that cones can offset some of the negative effects, or that birds adapt their reliance on cones to be more heavily dependent on them when other conditions are less favorable. That jays depend less on cones in areas of the SW with more monsoonal moisture suggests that these places potentially have more other foods available. Conversely, these are areas (like in NM) where there's been a lot of piñon die-off, so that may be related too.
